---
title: "STAT-S 610 Homework 1"
author: "Nahid Hasan Sagar"
date: "2025-09-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## STAT-S 610 Homework 1

# Loading necessary packages
```{r}
  library(ggplot2)
  library(wesanderson)
```

```{r}
  typeof(wes_palettes)
  typeof(names(wes_palettes))
  typeof(wes_palettes['AsteroidCity2'])
```
# Role of pal[1] and pal[2]
  So, in the code below pal is taking three color from wes_palettes and among those three color we are using two color to make a gradient on the following codes. Where pal[1] is the lower color middle one is while and higher color is pal[2]. So, pal[1] and pal[2] are used for setting lower and upper color in the gradient of a three color gradient.
```{r}
  data(heatmap)
    pal <- wes_palette("Rushmore", 3, type = "discrete")
    ggplot(heatmap, aes(x = X2, y = X1, fill = value)) +
        geom_tile() +
        scale_fill_gradient2(
            low = pal[1],
            mid = "white",
            high = pal[2],
        ) +
        scale_x_discrete(expand = c(0, 0)) +
        scale_y_discrete(expand = c(0, 0)) + coord_equal()
```
# Using for loop for the
```{r}
  for (i in 1:length(wes_palettes)) {
    pal <- wes_palette(names(wes_palettes)[i], 2, type = "discrete")
    print(ggplot(heatmap, aes(x = X2, y = X1, fill = value)) +
        geom_tile() +
        scale_fill_gradient2(
            low = pal[1],
            mid = "white",
            high = pal[2],
        ) +
        scale_x_discrete(expand = c(0, 0)) +
        scale_y_discrete(expand = c(0, 0)) + coord_equal())
  }
```
# With three different color
```{r}
  for (i in 1:length(wes_palettes)) {
    pal <- wes_palette(names(wes_palettes)[i], 3, type = "discrete")
    print(ggplot(heatmap, aes(x = X2, y = X1, fill = value)) +
        geom_tile() +
        scale_fill_gradient2(
            low = pal[1],
            mid = pal[2],
            high = pal[3],
        ) +
        scale_x_discrete(expand = c(0, 0)) +
        scale_y_discrete(expand = c(0, 0)) + coord_equal())
  }
```
# With one pair of color
```{r}
count <- 0
  for (i in 1:length(wes_palettes)) {
    pal <- wes_palette(names(wes_palettes)[i], 1, type = "discrete")
    for (j in 1:length(wes_palettes)) {
      if (i == j) {
        next
      }
      else {
        pal2 <- wes_palette(names(wes_palettes)[j], 1, type = "discrete")
        print(ggplot(heatmap, aes(x = X2, y = X1, fill = value)) + geom_tile() + scale_fill_gradient2( low = pal, mid = "White", high = pal2, ) + scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) + coord_equal())
        }
    }
  }
```